---
title: "Exemplary multistate model with non-exclusive categories"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simdat_analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

load(file = "../m_cmm_vignette.RData")


# and other packages that we need for estimation and visualization
library(survival)

```

This vignette aims to describe how one can analyze multistate frailty models
in which one target category (e.g., interacting with one's partner
and family members at the same time) consists of the combination of other categories.  

```{r setup}

# load the dena package with the data and functions
library(dena)

```

## load the example data 

```{r}
rm(list = ls())

data(simdat4) # see the end of the script for simulation code 


```

this dataset contains transition from the alone state to four different target states 
(1= family, 2= friend, 3= partner, 4= partner + family). This is a particular case of a
multistate model, where still all categories are competing against each other, but one
category (i.e., "partner + family") consists of the combination of two other categories. 

The ```type```variable in the dataset describes these target states: 

```{r}
head(simdat4)

table(simdat4$type)


```


## Estimating the multistate model

From a modeling perspective, there are no differences, compared to a standard application of the
multistate model, only that now there is a combined category of target states, of which the 
coefficients need to be interpreted in the light of the data collection process.


```{r, message = F}

m.cmm <- cmm(Surv(time, type) ~ Covariate1 + Covariate2 + (1 | id), dat = simdat4, verbose = T)

```

Here are the estimated fixed effects of the model:

```{r, message = F}

m.cmm[[1]]


```

or as a plot of the fixed effects:

```{r, message = F, warning = F, fig.width= 7, fig.height=7}
plot.cmm(m.cmm[[1]])

```

### Dataset simulation 

```{r, eval = F}


set.seed(921)
simdat4 <- frailtySurv::genfrail(N = 100, K = 50, 
                                 beta = c(log(1), -log(0.5)),
                                 frailty = "gamma", theta = 2, 
                                 censor.rate = 0,
                                 lambda_0=function(t, tau=4.6, C=0.01) (tau*(C*t)^tau)/t)

colnames(simdat4) <- c("id","rep","time","event","Covariate1","Covariate2")
simdat4$type <- factor(sample(c("partner", "family", "partner + family","friend"), nrow(simdat4), replace = T, prob = c(0.2,0.3,0.2,0.3)))

#save(simdat4, file = "../data/simdat4.RData")


```

